Acceleration of Global Search by Implementing Dual Estimates for Lipschitz Constant\thanks{This research was supported by the the Russian Science Foundation, project No.\,16-11-10150.}

### 2019-08-28 2019/NUMTA_Barkalov/LNCS/numta19.tex
Some major correction in Russian

%\usepackage[utf8x]{inputenc}
%\usepackage[english,russian]{babel}

@@

The paper considers global optimization problems of the form 
\begin{gather}
 \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}}, \label{problem}\\
 D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\} \label{D},
\end{gather}
where the objective function is a black-box function and it is assumed to satisfy the Lipschitz condition
\[
\left|\varphi(y_1)-\varphi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D,
\]
with the constant $L, \; L<\infty,$ unknown a priori.

%Russian
The assumption of the objective function to be Lipschitzian is typical for many approaches to the development of the global optimization algorithms \cite{Evtushenko2013,Zilinskas2010,Pinter1996,Strongin2000}. Moreover, the adaptive estimate of the unknown Lipschitz constant, based on the obtained search information, is one of the most important problems being solved in these algorithms. 
The value of the Lipschitz constant affects the convergence rate of the global optimization algorithms essentially. Therefore, the issue of its correct estimate is so important. 
The underestimation of the real value of this constant may result in losing the convergence of the algorithm to the global solution. At the same time, a too large value of the estimate of the constant could bring the objective function not matching its real behavior. This produces slow convergence of the algorithm to the global minimizer.

@@

\section{Global search algorithm and dimensionality reduction}
%Russian

@@

In the present paper, the approach based on the ideas of reduction of dimension with the use of the Peano-Hilbert curves \cite{Sergeyev2013,Strongin2000}, which continuously and unambiguously maps the unit interval $[0,1]$ onto the $N$-dimensional cube $D$ from (\ref{D}) has been used. By using this kind of mapping, it is possible to reduce the multidimensional problem (\ref{problem}) to a univariate problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}},
\]
where the function $\varphi(y(x))$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
with the H{\"o}lder constant $H$ linked to the Lipschitz constant $L$ by the relation
$ H=2 L \sqrt{N+3}$ and $y(x)$ is a Peano-Hilbert curve from $[0,1]$ onto $D$.

@@

 %If $\mu = 0$, set $\mu = 1$.
\item
For each interval $(x_{i-1}, x_i), \; 1\leq i \leq k,$  calculate the value 
$R(i)$ called the \textit{characteristic} of the interval
\begin{equation}\label{R}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r^2\mu^2\Delta_i}-2\frac{z_i+z_{i-1}-2z^*}{r\mu},
\end{equation}
where 
\begin{equation}\label{z}
z^*= \min_{0\leq i\leq k}z_i
\end{equation} 
and the real number $r>1$ is 
the input parameter of the algorithm.
\item 
Select the interval $(x_{t-1},x_t)$ corresponding to the maximum characteristic
\begin{equation}\label{MaxR}
R(t)= \max_{1 \leq i \leq k}R(i).
\end{equation}
\item
Carry out the next trial at the point $x^{k+1}\in(x_{t-1},x_t)$ calculated using
the following formula
\begin{equation}\label{xk1}
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}
\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N.
\end{equation}
\end{enumerate}

@@

%Russian
The global search algorithm presented in previous section is intended for solving the multiextremal problems, in which the objective function satisfies the Lipschitz condition. It is not necessary to define the value of the constant for the algorithm convergence. The estimation of the constant is performed in the course of global search on the base of available search information. 
According to the theorem from \cite{Strongin2000}, the sequence of the trial points $\{y^k\}$ will converge to the global minimizer $y^*$ if the condition 
\begin{equation}\label{cond}
r\mu > 2^{3-1/N}L\sqrt{N+3}.
\end{equation}
\noindent is satisfied. Thus,  an appropriate choice of the parameter $r$ from (\ref{R}) allows using the value $(r\mu) / (2^{3-1/N}\sqrt{N+3})$ as an estimate of the Lipschitz constant for the objective function $\varphi(y)$.

@@

The rules of the algorithm with two estimates of the Lipschitz constant reproduce the ones of global search algorithm completely except Rule 3 (the computation of the characteristic).

New rule of calculating the characteristic $R(i)$ of the interval $(x_{i-1}, x_i)$ will consist of the following operations:
\begin{itemize}
\item
Calculate the value $R_{glob}(i)$ corresponding to the larger estimate of the Lipschitz constant
\[
R_{glob}(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r_{glob}^2\mu^2\Delta_i}-2\frac{z_i+z_{i-1}-2z^*}{r_{glob}\mu}.
\]
\item
Calculate the value $R_{loc}(i)$ corresponding to the smaller estimate of the Lipschitz constant
\[
R_{loc}(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r_{loc}^2\mu^2\Delta_i}-2\frac{z_i+z_{i-1}-2z^*}{r_{loc}\mu}.
\]
\item
Determine the characteristic $R(i)$ as
\begin{equation}\label{pho}
R(i) = \max\{\rho R_{loc}(i),R_{glob}(i)\}, \textrm{where} \; \rho = \left(\frac{1-1/r_{glob}}{1-1/r_{loc}}\right)^2,
\end{equation}
and fix the value $r = r_{loc}$ if $\rho R_{loc}(i) > R_{glob}(i)$, otherwise fix $r=r_{glob}$.
\item
This value of $r$ is used in Rule 5 of the algorithm in the computing of the next trial point.   
\end{itemize}

@@

@@

%Russian
Each series of problems has been solved by the original global search algorithm (GSA) and by the method with two estimates of the Lipschitz constant (GSA-DL). The evolvent constructed using the parameter $m = 10$ was used for the dimensionality reduction in both algorithms. The relative accuracy of the solution search was $\delta = 0.01$.% for $N=3,4$ and $\delta = 0.02$ for $N=5$. 
The maximum allowable number of iterations per problem was $K_{max} = 10^6$.

@@

The lower curves in Fig. \ref{oper4} and \ref{oper5} feature GSA whereas the upper ones --- GSA-DL. Such positions of the curves show the algorithm with two estimates of the Lipschitz constant provides much faster solving of the problem series in average that the algorithm using single estimate of the constant.

>>>>

\usepackage[utf8x]{inputenc}
\usepackage[english,russian]{babel}

@@

The paper considers global optimization problems of the form 
\begin{gather}
 \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}}, \label{problem}\\
 D=\left\{y\in R^N: a_i\leq y_i \leq b_i, \; a_i,b_i\in R, \;  1\leq i \leq N\right\} \label{D},
\end{gather}
where the objective function is a black-box function and it is assumed to satisfy the Lipschitz condition
\[
\left|\varphi(y_1)-\varphi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D,
\]
with the constant $L, \; L<\infty,$ unknown a priori.

The assumption of the objective function to be Lipschitzian is typical for many approaches to the development of the global optimization algorithms \cite{Evtushenko2013,Zilinskas2010,Pinter1996,Strongin2000}. Moreover, the adaptive estimate of the unknown Lipschitz constant, based on the obtained search information, is one of the most important problems being solved in these algorithms. 
The value of the Lipschitz constant affects the convergence rate of the global optimization algorithms essentially. Therefore, the issue of its correct estimate is so important. 
The underestimation of the real value of this constant may result in losing the convergence of the algorithm to the global solution. At the same time, a too large value of the estimate of the constant could bring the objective function not matching its real behavior. This produces slow convergence of the algorithm to the global minimizer.

@@

\section{Global search algorithm and dimensionality reduction}

@@

In the present paper, the approach based on the ideas of reduction of dimension with the use of the Peano-Hilbert curves \cite{Sergeyev2013,Strongin2000}, which continuously and unambiguously maps the unit interval $[0,1]$ onto the $N$-dimensional cube $D$ from (\ref{D}) has been used. By using this kind of mapping, it is possible to reduce the multidimensional problem (\ref{problem}) to a univariate problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}},
\]
where the function $\varphi(y(x))$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
with the H{\"o}lder constant $H$ linked to the Lipschitz constant $L$ by the relation
$ H=2 L \sqrt{N+3}$ and $y(x)$ is a Peano-Hilbert curve from $[0,1]$ onto $D$.
%Russian
Отметим, что теоретически Peano-Hilbert curve $y(x)$ определяется as a limit object. Поэтому при практической реализации может быть построено лишь некоторое приближение к истинной space-filling curve. Методы построения подобных approximations (называемых evolvents) рассмотрены в \cite{Sergeyev2013,Strongin2000}. При этом точность приближения развертки к истинной кривой $y(x)$ зависит от плотности развертки $m$ (являющейся параметром ее построения) и составляет величину порядка $2^{-m}$ по каждой координате.

@@

\item
For each interval $(x_{i-1}, x_i), \; 1\leq i \leq k,$  calculate the value 
$R(i)$ called the \textit{characteristic} of the interval
\begin{equation}\label{R}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r^2\mu^2\Delta_i}-2\frac{z_i+z_{i-1}-2z^*}{r\mu},
\end{equation}
where 
\begin{equation}\label{z}
z^*= \min_{0\leq i\leq k}z_i
\end{equation} 
and the real number $r>1$ is 
the input parameter of the algorithm.
\item 
Select the interval $(x_{t-1},x_t)$ corresponding to the maximum characteristic
\begin{equation}\label{MaxR}
R(t)= \max_{1 \leq i \leq k}R(i).
\end{equation}
\item
Carry out the next trial at the point $x^{k+1}\in(x_{t-1},x_t)$ calculated using
the following formula
\begin{equation}\label{xk1}
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}
\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N.
\end{equation}
\end{enumerate}

@@

The global search algorithm presented in previous section is intended for solving the multiextremal problems, in which the objective function satisfies the Lipschitz condition. It is not necessary to define the value of the constant for the algorithm convergence. The estimation of the constant is performed in the course of global search on the base of available search information. 
According to the theorem from \cite{Strongin2000}, the sequence of the trial points $\{y^k\}$ will converge to the global minimizer $y^*$ if the condition 
\begin{equation}\label{cond}
r\mu > 2^{3-1/N}L\sqrt{N+3}.
\end{equation}
\noindent is satisfied. Thus,  an appropriate choice of the parameter $r$ from (\ref{R}) allows using the value $(r\mu) / (2^{3-1/N}\sqrt{N+3})$ as an estimate of the Lipschitz constant for the objective function $\varphi(y)$.

@@

%Russian
The rules of the algorithm with two estimates of the Lipschitz constant reproduce the ones of global search algorithm completely except Rule 3 (the computation of the characteristic) and Rule 4 (поиск интервала с максимальной характеристикой).

New rule of calculating the characteristic $R(i)$ of the interval $(x_{i-1}, x_i)$ will consist of the following operations:
\begin{itemize}
\item
Calculate the value $R_{glob}(i)$ corresponding to the larger estimate of the Lipschitz constant
\[
R_{glob}(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r_{glob}^2\mu^2\Delta_i}-2\frac{z_i+z_{i-1}-2z^*}{r_{glob}\mu}.
\]
\item
Calculate the value $R_{loc}(i)$ corresponding to the smaller estimate of the Lipschitz constant
\[
R_{loc}(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r_{loc}^2\mu^2\Delta_i}-2\frac{z_i+z_{i-1}-2z^*}{r_{loc}\mu}.
\]
\item
Determine the characteristic $R(i)$ as
\begin{equation}\label{pho}
R(i) = \max\{\rho R_{loc}(i),R_{glob}(i)\}, \textrm{where} \; \rho = \left(\frac{1-1/r_{glob}}{1-1/r_{loc}}\right)^2,
\end{equation}   
\end{itemize}

%Russian
Новые правила поиска интервала с максимальной характеристикой будут состоять в следующем:
\begin{itemize}
\item 
Select the interval $(x_{t-1},x_t)$ corresponding to the maximum characteristic
\[
R(t)= \max_{1 \leq i \leq k}R(i).
\]
\item
Fix the value $r = r_{loc}$ if $\rho R_{loc}(t) > R_{glob}(t)$, otherwise fix $r=r_{glob}$.
\item
This value of $r$ is used in Rule 5 of the algorithm in the computing of the next trial point.
\end{itemize}

@@

%Russian
Численное сравление алгоритмов было выполнено посредством решения нескольких серий задач, порожденных генератором GKLS \cite{Gaviano2003}. Данный генератор многоэкстремальных функций широко используется для сравнения методов глобальной оптимизации (see, for example, \cite{Barkalov2018,Paulavicius2014,Sergeyev2015}). В данном исследовании были решены шесть серий по 100 задач размерностей $N = 3,4,5$. Для каждой размерности генерировались \textit{Simple} and \textit{Hard} problems, отличающиеся размерами attraction region локальных и глобального minimizer.
Задача считалась решенной, если метод проводил очередное испытание в точке $y^k$, которая находилась в $\delta$-октрестности global minimizer $y^\ast$, 
т.е. $\left\|y^k-y^\ast\right\| <\delta\left\|b-a\right\|$, где  $a$ and $b$ --- границы search domain $D$.

@@

Each series of problems has been solved by the original global search algorithm (GSA) and by the method with two estimates of the Lipschitz constant (GSA-DL). The evolvent constructed using the parameter $m = 10$ was used for the dimensionality reduction in both algorithms. The relative accuracy of the solution search was $\delta = 0.01$. 
The maximum allowable number of iterations per problem was $K_{max} = 10^6$.

@@

The lower curves in Fig. \ref{oper4} and \ref{oper5} feature GSA whereas the upper ones --- GSA-DL. Such positions of the curves show the algorithm with two estimates of the Lipschitz constant provides much faster solving of the problem series in average that the algorithm using single estimate of the constant.
%Russian
Отметим, что для решения задач всех классов (кроме класса  \textit{Hard} при $N=5$) метод GSA-DL требует примерно в два раза меньше испытаний, чем GSA. Ухудшение результатов на классе \textit{Hard} при $N=5$ объясняется сложностью функций данного класса, характеризующихся большими attraction region локальных минимумов и узкой attraction region глобального минимума. Для корректного решения таких задач метод  GSA-DL чаще использует бОльшую оценку константы Липшица, что сближает скорость работы алгоритмов GSA и GSA-DL.
